Supplementary material for: non-coding RNAs of bird genomes

### 2014-05-19 paper/manuscript.tex
More responses to referees, tidying, etc.

%\title{Non-coding RNAs of bird genomes}
%\title{Wins and losses: Non-coding RNAs of bird genomes}
\title{Conservation and losses of avian non-coding RNAs}

@@

\begin{abstract}
Here we present the results of a large-scale bioinformatic annotation
of non-coding RNAs in 48 avian genomes. Our approach uses
probabilistic models of hand-curated families from the Rfam database
to infer conserved RNA families within each avian genome. We
supplement these annotations with predictions from the tRNA annotation
tool, tRNAscan-SE and microRNAs from miRBase.  We show a significant
number of lncRNAs are surprisingly well conserved between birds and
mammals including several intriguing cases where the reported
mammalian lncRNA function is not conserved in birds.  We also
demonstrate extensive conservation of classical ncRNAs (e.g., tRNAs)
and more recently discovered ncRNAs (e.g., snoRNAs and miRNAs) in
birds. Furthermore, we describe numerous ``losses'' of several RNA
families, and attribute these to genuine loss, divergence or missing
data.  In particular, issues with assembling microchromosomes using
next-generation sequencing methods. These combined results illustrate
the utility of applying homology based methods for annotating novel
vertebrate genomes and illustrate many complex evolutionary patterns
within the avian ncRNA cohort.
\end{abstract}

@@

Independent benchmarks of bioinformatic annotation tools have shown
that the CM approaches dramatically out-perform alternative methods
\cite{Freyhult:2007}, although its sensitivity is limited for the most
rapidly evolving families such as vault RNAs or telomerase RNA
\cite{Menzel:09a}.

@@

The publication of 48 avian genomes, including the previously
published chicken
\cite{International_Chicken_Genome_Sequencing_Consortium:2004}, zebra
finch \cite{Warren:2010} and turkey \cite{Dalloul:2010} with the
recently published 45 avian genomes
\cite{birds:14,birds:14a,Huang:2013,Zhan:2013,Shapiro:2013,Howard:2013,Li:2014},
provides exciting opportunities to explore ncRNA conservation in
unprecedented detail.
%{\bf MORE ON BIRD GENOMES. ESP. MACRO \& MICRO
%  CHROMOSOMES.}

In the following we explore the conservation patterns of the major
classes of avian ncRNA in further detail.  Using homology search tools
and evolutionary constraints, we produce an accurate set of genome
annotations for 48 predominantly non-model bird species for ncRNAs
that are conserved across the avian species. This conservative set of
annotations is expected to contain the core avian ncRNAs. We focus our
report on the unusual results within the avian lineages. These are
either unexpectedly well-conserved ncRNAs or unexpectedly
poorly-conserved ncRNAs. The former are ncRNAs that were not expected
to be conserved between the birds and the other vertebrates,
particularly those ncRNAs whose function is not conserved in birds.
The latter are apparent losses of ncRNAs expected to be conserved;
Here, we consider three categories of such ``loss'': First, genuine
gene losses in the avian lineage where ncRNAs well conserved in other
vertebrates are completely absent in birds.  Second, ``divergence''
where ncRNAs have undergone such significant sequence and structural
alternations that homology search tools can no longer detect a
relationship between other vertebrate exemplars and avian varieties.
Third, ``missing'' ncRNAs that failed to be captured in the available,
largely fragmented, avian genomes.  We postulate that the latter
category is likely to be prevalent in comparative avian genome studies
given the distinctive organisation of the avian genome. Namely, the
avian karyotype is characterised by a large number of chromosomes
(average $2n \approx 80$) generally consisting of a few larger
``macrochromosomes'' and many smaller ``microchromosomes''
\cite{Griffin:2007}. This ’so many, so small’ pattern presents
significant assembly challenges \cite{Ellegren:2005}. Indeed, of the
48 published avian genomes, 20 of which are high-coverage ($>50X$ ),
only two are chromosomally assembled (chicken and zebra finch;
\cite{birds:14,Warren:2010}).

@@

There is substantial gain and loss of lncRNA
\cite{Cabili:2011,Kutter:2012} and more generally ncRNA loci over
evolutionary time \cite{Hoeppner:2012}. It is difficult to assess how
much of this is due to bioinformatic sequence alignment tools failing
to align correctly below 60-50\% sequence identity
\cite{Gardner:2005:Nucleic-Acids-Res:15860779} or due to genuine gains
and losses. Nevertheless, sequence conservation, generally speaking,
provides useful evidence for gene and function conservation.

We have identified 66,879 loci in 48 avian genomes that share sequence
similarity with previously characterised ncRNAs and are conserved in
$>10\%$ of these avian genomes. These loci have been classified into
626 different families, the majority of which are miRNAs and snoRNAs
(summarised in Table~\ref{table:1}). Out of necessity we have selected
the most charismatic families for further discussion. These are
lncRNAs that appear to be conserved between mammals and aves and the
cases of apparent loss of genes that conserved in most other
vertebrates.

@@

The bulk of the ``unusually well conserved RNAs'' belong to the long
non-coding RNA (lncRNA) group.  The lncRNAs are a diverse group of
RNAs that have been implicated in a multitude of functional processes
\cite{Rinn:2007,Chow:2005,Guttman:2009,Ulitsky:2013}. These RNAs have largely been
characterised in mammalian species, particularly human and
mouse. Consequently, we generally do not expect these to be conserved
outside of mammalia. Notable examples include Xist \cite{Duret:2006}
and H19 \cite{Smits:2008}.  There is emerging evidence for the
conservation of ``mammalian'' lncRNAs in other vertebrates
\cite{Chodroff:2010,Ulitsky:2011}), however, like most lncRNAs, the
function of these lncRNAs remains largely unknown. Here, we show the
conservation of several well-characterised lncRNAs of known function
in humans.

The CM based approach is appropriate for most classes of ncRNA, but
the long non-coding RNAs (lncRNAs) are a particular challenge
\cite{Guttman:2009}. 
CMs cannot model the exon-intron structures of spliced lncRNAs, nor do
they deal elegantly with the repeats that many lncRNAs
host. Consequently in the latest release of Rfam the lncRNA families
that were added were composed of local conserved (and possibly
structured elements) within lncRNAs, analogous to the ``domains''
housed within protein sequences \cite{Burge:2013}. Whilst some these
regions may not reflect functional RNA elements but instead regulatory
regions, enhancers or insulators, their syntenic conservation still
provides an indication of lncRNA conservation
\cite{diederichs2014four}.

@@

The HOX cluster lncRNAs HOTAIRM1 (5 RNA-domains), HOXA11-AS1 (6
RNA-domains), and HOTTIP (4 RNA domains) are remarkably well conserved. In
the human genome they are located in the HOXA cluster (hg coordinates
chr7:27135743-27245922), one of the most highly conserved regions in
vertebrate genomes \cite{PascualAnaya:13}, in antisense orientation between
HoxA1 and HoxA2, between HoxA11 and HoxA13, and upstream of HoxA13,
respectively. Conservation and expression of HOTAIRM1 and HOXA11-AS1 within
the HOXA cluster has been studied in some detail in marsupials
\cite{Yu:12}.  Of the 15 RNA-domains five and six representing all three
lncRNAs were recovered in the alligator and turtle genomes. All of them
appear in the correct order at the expected, syntenically conserved
positions within the HOXA cluster.  In the birds, where two or more of the
HOX cluster lncRNA RNA-domains were predicted on the same scaffold, this
gene order and location within HOX was also preserved.

@@

The NBR2 lncRNA and BRCA1 gene share a bidirectional promotor
\cite{Xu:1997}. Both are expressed in a broad range of
tissues. Extensive research on BRCA1 has shown that it is involved in
DNA repair \cite{Moynahan:1999}. The function of NBR2 remains unknown,
yet its conservation across the vertebrates certainly implies a
function (See Figure~\ref{fig:1}).

@@

Is it possible that the avian annotations also contains excessive
pseudogenes. However, it has previously been noted that avian genomes
are significantly smaller than other vertebrate species
\cite{International_Chicken_Genome_Sequencing_Consortium:2004}. We
have also noted a corresponding reduction in the number of paralogs
and presumed ncRNA-derived pseudogenes in the avian genomes (See
Supplemental Figure~12). The problematic human families, U6 snRNA, SRP
RNA and Y RNA have, for example, just 26, 4 and 3 annotations
respectively in the chicken genome and 13, 3 and 3 annotations
respectively, on average, in the 48 avian genomes used here.
Therefore, we suspect that the majority of our annotations are in fact
functional orthologs.

@@

@@

Chicken ncRNA predictions were validated using two separate RNA-seq
data sets. The first data set (Bioproject PRJNA204941) contains 971
million reads and comprises 27 samples from 14 different chicken
tissues sequenced on Illumina HiSeq2000 using a small RNA-seq
protocol. The second data set (personal commun. Matt Schwartz and Igor
Ulitsky) contains 1,46 billion Illumina HiSeq reads sequenced from
whole chicken embryo RNA from 7 stages using a strand-specific dUTP
protocol. Raw reads were checked for quality and adapters clipped if
required by the protocol. Preprocessed reads were mapped to the
galGal4 reference genome using SEGEMEHL short read aligner
\cite{segemehl} and then overlapped with the ncRNA annotations.

@@

Heatmaps showing the presence/absence and approximate genomic copy
number of ``unusually, well conserved RNAs'' (particularly the
lncRNAs) on the left and families that have been identified as
surprising RNA Losses, divergence or missing data. In several cases
functionally related families have also been included, e.g. the RNA
components of the major and minor spliceosomes: U1, U2, U4, U5 and U6;
and U11, U12, U4atac, U5 and U6atac, respectively.\label{fig:1}

>>>>

%\title{Non-coding RNAs of bird genomes}
%\title{Wins and losses: Non-coding RNAs of bird genomes}
\title{Conservation and losses of avian non-coding RNA loci}

@@

\begin{abstract}
Here we present the results of a large-scale bioinformatic annotation
of non-coding RNA loci in 48 avian genomes. Our approach uses
probabilistic models of hand-curated families from the Rfam database
to infer conserved RNA families within each avian genome. We
supplement these annotations with predictions from the tRNA annotation
tool, tRNAscan-SE and microRNAs from miRBase.  We show a number of
lncRNA associated loci are conserved between birds and mammals,
including several intriguing cases where the reported mammalian lncRNA
function is not conserved in birds.  We also demonstrate extensive
conservation of classical ncRNAs (e.g., tRNAs) and more recently
discovered ncRNAs (e.g., snoRNAs and miRNAs) in birds. Furthermore, we
describe numerous ``losses'' of several RNA families, and attribute
these to genuine loss, divergence or missing data.  In particular, we
show that many of these losses are due to issues with assembling
microchromosomes using next-generation sequencing methods. These
combined results illustrate the utility of applying homology-based
methods for annotating novel vertebrate genomes.
\end{abstract}

@@

Independent benchmarks of bioinformatic annotation tools have shown
that the CM approaches out-perform alternative methods
\cite{Freyhult:2007}, although its sensitivity is limited for the most
rapidly evolving families such as vault RNAs or telomerase RNA
\cite{Menzel:09a}.

@@

The publication of 48 avian genomes, including the previously
published chicken
\cite{International_Chicken_Genome_Sequencing_Consortium:2004}, zebra
finch \cite{Warren:2010} and turkey \cite{Dalloul:2010} with the
recently published 45 avian genomes
\cite{birds:14,birds:14a,Huang:2013,Zhan:2013,Shapiro:2013,Howard:2013,Li:2014},
provides an exciting opportunity to explore ncRNA conservation in
unprecedented detail.
%{\bf MORE ON BIRD GENOMES. ESP. MACRO \& MICRO
%  CHROMOSOMES.}

In the following we explore the conservation patterns of the major
classes of avian ncRNA loci in further detail.  Using homology search
tools and evolutionary constraints, we have produced a set of genome
annotations for 48 predominantly non-model bird species for ncRNAs
that are conserved across the avian species. This conservative set of
annotations is expected to contain the core avian ncRNA loci. We focus
our report on the unusual results within the avian lineages. These are
either unexpectedly well-conserved ncRNAs or unexpectedly
poorly-conserved ncRNAs. The former are ncRNA loci that were not
expected to be conserved between the birds and the other vertebrates,
particularly those ncRNAs whose function is not conserved in birds.
The latter are apparent losses of ncRNA loci expected to be conserved;
Here, we consider three categories of such ``loss'': First, genuine
gene losses in the avian lineage where ncRNAs well conserved in other
vertebrates are completely absent in birds.  Second, ``divergence''
where ncRNAs have undergone such significant sequence and structural
alternations that homology search tools can no longer detect a
relationship between other vertebrate exemplars and avian varieties.
Third, ``missing'' ncRNAs that failed to be captured in the available,
largely fragmented, avian genomes.  We postulate that the latter
category is likely to be prevalent in comparative avian genome studies
given the distinctive organisation of the avian genome. Namely, the
avian karyotype is characterised by a large number of chromosomes
(average $2n \approx 80$) generally consisting of approximately 5
larger ``macrochromosomes'' and many smaller ``microchromosomes''
\cite{Griffin:2007}. This ’so many, so small’ pattern presents
significant assembly challenges \cite{Ellegren:2005}. Indeed, of the
48 published avian genomes, 20 of which are high-coverage ($>50X$),
only two are chromosomally assembled (chicken and zebra finch;
\cite{birds:14,Warren:2010}).

@@

There is substantial gain and loss of lncRNAs
\cite{Cabili:2011,Kutter:2012} and more generally ncRNA loci over
evolutionary time \cite{Hoeppner:2012}. It is difficult to assess how
many of these observations are due to limited bioinformatic sequence
alignment tools (these generally fail align correctly below 60-50\%
sequence identity \cite{Gardner:2005:Nucleic-Acids-Res:15860779}) or
due to genuine gains and losses. Nevertheless, sequence conservation,
generally speaking, provides useful evidence for gene and function
conservation.

We have identified 66,879 loci in 48 avian genomes that share sequence
similarity with previously characterised ncRNAs and are conserved in
$>10\%$ of these avian genomes. These loci have been classified into
626 different families, the majority of which correspond to miRNAs and
snoRNAs (summarised in Table~\ref{table:1}). Out of necessity we have
selected the most charismatic families for further discussion. These
include the lncRNAs that appear to be conserved between Mammals and
Aves and the cases of apparent loss of genes that conserved in most
other Vertebrates.

@@

The bulk of the ``unusually well conserved RNAs'' belong to the long
non-coding RNA (lncRNA) group.  The lncRNAs are a diverse group of
RNAs that have been implicated in a multitude of functional processes
\cite{Rinn:2007,Chow:2005,Guttman:2009,Ulitsky:2013}. These RNAs have largely been
characterised in mammalian species, particularly human and
mouse. Consequently, we generally do not expect these to be conserved
outside of mammalia. Notable examples include Xist \cite{Duret:2006}
and H19 \cite{Smits:2008}.  There is emerging evidence for the
conservation of ``Mammalian'' lncRNAs in other Vertebrates
\cite{Chodroff:2010,Ulitsky:2011}), however, like most lncRNAs, the
function of these lncRNAs remains largely unknown. Here, we show the
conservation of several well-characterised lncRNAs of known function
in humans.

The CM based approach is appropriate for most classes of ncRNA, but
the lncRNAs are a particular challenge \cite{Guttman:2009}.  CMs
cannot model the exon-intron structures of spliced lncRNAs, nor do
they deal elegantly with the repeats that many lncRNAs
host. Consequently in the latest release of Rfam the lncRNA families
that were added were composed of local conserved (and possibly
structured elements) within lncRNAs, analogous to the ``domains''
housed within protein sequences \cite{Burge:2013}. Whilst some these
regions may not reflect functional RNA elements but instead regulatory
regions, enhancers or insulators, their syntenic conservation still
provides an indication of lncRNA conservation
\cite{diederichs2014four}.

@@

The HOX cluster lncRNAs HOTAIRM1 (5 RNA-domains), HOXA11-AS1 (6
RNA-domains), and HOTTIP (4 RNA domains) are conserved across the
Mammalian and Avian lineages. In the human genome they are located in
the HOXA cluster (hg coordinates chr7:27135743-27245922), one of the
most highly conserved regions in vertebrate genomes
\cite{PascualAnaya:13}, in antisense orientation between HoxA1 and
HoxA2, between HoxA11 and HoxA13, and upstream of HoxA13,
respectively. Conservation and expression of HOTAIRM1 and HOXA11-AS1
within the HOXA cluster has been studied in some detail in marsupials
\cite{Yu:12}.  Of the 15 RNA-domains five and six representing all
three lncRNAs were recovered in the alligator and turtle genomes. All
of them appear in the correct order at the expected, syntenically
conserved positions within the HOXA cluster.  In the birds, where two
or more of the HOX cluster lncRNA RNA-domains were predicted on the
same scaffold, this gene order and location within HOX was also
preserved.

@@

The NBR2 lncRNA and BRCA1 gene share a bidirectional promotor
\cite{Xu:1997}. Both are expressed in a broad range of
tissues. Extensive research on BRCA1 has shown that it is involved in
DNA repair \cite{Moynahan:1999}. The function of NBR2 remains unknown,
yet its conservation across the vertebrates certainly implies a
function (See Figure~\ref{fig:1}). Possibly at the DNA level, however,
function at the RNA level cannot be ruled out at this stage.

@@

It is possible that the avian annotations also contains excessive
pseudogenes. However, it has previously been noted that avian genomes
are significantly smaller than other vertebrate species
\cite{International_Chicken_Genome_Sequencing_Consortium:2004}. We
have also noted a corresponding reduction in the number of paralogs
and presumed ncRNA-derived pseudogenes in the avian genomes (See
Supplemental Figure~12). The problematic human families, U6 snRNA, SRP
RNA and Y RNA have, for example, just 26, 4 and 3 annotations
respectively in the chicken genome and 13, 3 and 3 annotations
respectively, on average, in the 48 avian genomes used here.
Therefore, we conclude that the majority of our annotations are in fact
functional orthologs.

@@

The 48 bird genome sequences used for the following analyses are available from
the phylogenomics analysis of birds website \cite{birdphylogen}.

@@

Chicken ncRNA predictions were validated using two separate RNA-seq
data sets. The first data set (Bioproject PRJNA204941) contains 971
million reads and comprises 27 samples from 14 different chicken
tissues sequenced on Illumina HiSeq2000 using a small RNA-seq
protocol. The second data set (SRA accession SRP041863) contains 1,46
billion Illumina HiSeq reads sequenced from whole chicken embryo RNA
from 7 stages using a strand-specific dUTP protocol. Raw reads were
checked for quality and adapters clipped if required by the
protocol. Preprocessed reads were mapped to the galGal4 reference
genome using SEGEMEHL short read aligner \cite{segemehl} and then
overlapped with the ncRNA annotations.

@@

Heatmaps showing the presence/absence and approximate genomic copy
number of ``unusually, well conserved RNAs'' (particularly the
lncRNAs) on the left and families that have been identified as
RNA losses, divergence or missing data. In several cases
functionally related families have also been included, e.g. the RNA
components of the major and minor spliceosomes: U1, U2, U4, U5 and U6;
and U11, U12, U4atac, U5 and U6atac, respectively.\label{fig:1}

### 2014-05-19 paper/supplementary-material.tex
More responses to referees, tidying, etc.

\begin{figure}[ht]
  \includegraphics[width=0.95\textwidth]{figures/Y.pdf}
  \caption[]{...}\label{fig:6}
\end{figure}

@@

\begin{figure}[ht]
\includegraphics[width=0.99\textwidth]{/home/ppg15/people/birds/mario/chr1-147253243-147253923-miRNA-cluster.pdf}
  \caption[]{microRNA cluster: mir-19\_3, mir-20\_2, mir-19\_2, mir-18\_2 \& mir-17\_1.}\label{fig:10}
\end{figure}

\begin{figure}[ht]
\includegraphics[width=0.99\textwidth]{/home/ppg15/people/birds/mario/chr4-3946041-3946705-miRNA-cluster.pdf}
  \caption[]{microRNA cluster: mir-92\_2, mir-19\_1, mir-20\_1, mir-18\_1, mir-106\_1.}\label{fig:11}
\end{figure}

\begin{figure}[ht]
\includegraphics[width=0.99\textwidth]{/home/ppg15/people/birds/mario/chr23-2507395-2508461-snoRNA-cluster.pdf}
  \caption[]{H/ACA box snoRNA cluster: SNORA61, SNORA44, SNORA16.}\label{fig:12}
\end{figure}

\begin{figure}[ht]
\includegraphics[width=0.99\textwidth]{/home/ppg15/people/birds/mario/chr27-4040511-4042052-tRNA-cluster.pdf}
  \caption[]{Cysteine tRNA cluster.}\label{fig:13}
\end{figure}

\begin{figure}[ht]
\includegraphics[width=0.475\textwidth]{figures/high-copy-numbers.pdf}
\includegraphics[width=0.475\textwidth]{figures/high-copy-numbers-U6-SRP.pdf}
  \caption[]{{\bf Left:} A comparison of the genomic copy numbers for
    26 of the most abundant RNAs found in human, reptile and the bird
    genomes. {\bf Right:} Bit-score distributions for the Rfam U6,
    Metazoan SRP and the Y RNA covariance models for human and chicken
    paralogues/pseudogenes. }\label{fig:14}
\end{figure}

>>>>

\begin{figure}[ht]
  \includegraphics[width=0.95\textwidth]{figures/Y.pdf}
  \caption[]{Syntenic conservation of the Y RNA cluster in avian and
    reptile genomes.}\label{fig:6}
\end{figure}

@@

\begin{figure}[ht]
%\includegraphics[width=0.99\textwidth]{/home/ppg15/people/birds/mario/chr1-147253243-147253923-miRNA-cluster.pdf}
\includegraphics[width=0.99\textwidth]{figures/mir-17-I-1-cluster.pdf}
  \caption[]{microRNA 17 I-1 cluster: mir-92\_1, mir-19\_3, mir-20\_2,
    mir-19\_2, mir-18\_2 \& mir-17\_1. The figure indicates the
    genomic location and the RNA-seq read-depths for these 6
    microRNAs.}\label{fig:10}
\end{figure}

\begin{figure}[ht]
\includegraphics[width=0.99\textwidth]{figures/mir-17-I-X-cluster.pdf}
  \caption[]{microRNA 17 I-X cluster: mir-363\_1, mir-92\_2,
    mir-19\_1, mir-20\_1, mir-18\_1, mir-106\_1. The figure indicates
    the genomic location and the RNA-seq read-depths for these 6
    microRNAs.}\label{fig:11}
\end{figure}

\begin{figure}[ht]
\includegraphics[width=0.99\textwidth]{figures/chr23-2507395-2508461-snoRNA-cluster.pdf}
  \caption[]{H/ACA box snoRNA cluster: SNORA61, SNORA44, SNORA16. The figure indicates
    the genomic location and the RNA-seq read-depths for these 3
    snoRNAs.}\label{fig:12}
\end{figure}

\begin{figure}[ht]
\includegraphics[width=0.99\textwidth]{figures/chr27-4040511-4042052-tRNA-cluster.pdf}
  \caption[]{Cysteine tRNA cluster. The figure indicates
    the genomic location and the RNA-seq read-depths for these 5
    transfer RNAs.}\label{fig:13}
\end{figure}

\begin{figure}[ht]
\includegraphics[width=0.475\textwidth]{figures/high-copy-numbers.pdf}
\includegraphics[width=0.475\textwidth]{figures/high-copy-numbers-U6-SRP.pdf}
  \caption[]{{\bf Left:} A comparison of the genomic copy numbers for
    26 of the most abundant RNAs found in human, reptile and the bird
    genomes. {\bf Right:} Bit-score distributions for the Rfam U6,
    Metazoan SRP and the Y RNA covariance models for human and chicken
    paralogues/pseudogenes. Loci with high bit-scores are more likely
    to be functional than those with low bit-scores (due to excess
    variation that conflicts with the canonical RNA
    models).}\label{fig:14}
\end{figure}
